---
title: "Graduation with DemoTools"
author: "José Manuel Aburto, Ilya Kashnitsky, Monica Alexander, Jorge Cimentada, Tim Riffe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Graduation with DemoTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, 
  fig.height = 6,
  fig.align = "center"
)

```

## How to use `DemoTools` to graduate counts in grouped ages

### What is graduation?
Graduation is a practice used to derive figures for _n_-year age groups, for example 5-year age groups from census data, that are corrected for net reporting error [@siegel2004methods]. The basic idea is to fit different curves to the original  _n_-year and redistribute them into single-year values. These techniques are designed so that the sum of the interpolated single-year values is consistent with the total for the groups as a whole. Among the major graduation methods are the Sprague [@sprague1880explanation] and Beers [@beers1945modified] oscilatory methods, monotone spline and the uniform distribution. More recently, the penalized composite link model (pclm) was proposed to ungroup coarsely aggregated data [@rizzi2015efficient]. All of these methodologies are implemented in `DemoTools`.

### Why graduate?
One of the main purposes of graduation is to refine the detail of available data in published statistics or surveys. This allows to estimate numbers of persons in single years of age from data originally in 5-year age groups. In practice, graduation of mortality curves has been very important to estimate reliable life tables, but also in fertility studies graduation is useful for the analysis of data from populations where demographic records are defective [@brass1960graduation].

### How is this different from smoothing?
The main difference between graduation and smoothing is that graduation redistributes the counts or events into single-year age groups with the constrain that they sum to the original total. The selection of the method to be used depends mostly on the balance between smoothness and closeness of fit to the data. Therefore, the analyst must decide which weight should be given to these two opposing considerations. 

## Read in data
```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
library(DemoTools)
# 'pop1m_ind' available as package data
Value <- pop5_mat[,4]
Age   <- as.integer(rownames(pop5_mat))
  
plot(Age, Value, type = 'p',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Original 5-year age-group population')
```

## Methods choices

### Sprague
@sprague1880explanation suggested the fifth-difference method based on one equation that depends on two polynomials of degree four forced to have a common ordinate, tangent and radius of curvature at $Y_{n+2}$ and at $Y_{n+3}$:

\begin{equation*}
y_{n+2+x} = \frac{(x+2)}{1!}\Delta y_n +
\frac{(x+2)(x+1)}{2!}\Delta y^2_n +
\frac{(x+2)(x+1)x}{3!}\Delta y^3_n +
\frac{(x+2)(x+1)(x-1)x}{4!}\Delta y^4_n +
\frac{x^3(x-1)(5x-7)}{4!}\Delta y^5_n 
\end{equation*}

There are six observations $Y_{n},Y_{n+1},Y_{n+2},Y_{n+3},Y_{n+4}$ and $Y_{n+5}$ involved in the leading differences, $\Delta y_{n},\dots,\Delta^5 y_n $. The interpolated points fall on curves that pass through the given points, and the interpolated subdivisions add up to the data in the original groups. To implement this method in `DemoTools`, use the function `graduate` with the option for method specified as `sprague`. 


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
sprague <- graduate(Value, Age, method = "sprague")
single.age  <- names2age(sprague)

plot(single.age, sprague, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with Sprague method')

#check the totals
sum(sprague)
sum(Value)
```
### Beers

Following a similar idea, Beers interpolated two overlapping curves minimizing the squares of the differences within the interpolation range [@beers1945modified]. Specifically, Beers did this by minimizing fifth differences for a six-term formula, refered to as the 'Ordinary' Beers method [@beers1945six]. Subsequently, the ordinay formula was modified to relax the requirement that the given value be reproduced and yield smoother interpolated results, refered to as 'Modified' Beers method [@beers1945modified].

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
beers.ordinary <- graduate(Value, Age, method = "beers(ord)")
beers.modified <- graduate(Value, Age, method = "beers(mod)")

single.age  <- names2age(beers.ordinary)
plot(single.age, beers.ordinary, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with Beers methods')
lines(single.age,beers.modified, col = "red")
legend("topright",lty=1,col=c(1,2),legend=c("ordinary (5-yr constrained)",
                                            "modified (smoother)"))
#check the totals
sum(beers.ordinary)
sum(beers.modified)
sum(Value)
```

### Monotone spline
Graduation by a monotone spline is another option available in `DemoTools`. This option offers an algorithm to produce interpolants that preserve properties such as monotonicity or convexity that are present in the data [@fritsch1980monotone]. These are desirable conditions to not introduce biased details that cannot be ascertained from the data [@hyman1983accurate]. To run this algorithm in `DemoTools` the option `mono` should be selected as follows:

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
mono.grad <- graduate(Value, Age, method = "mono")

single.age  <- names2age(mono.grad)
plot(single.age, mono.grad, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with monotone spline')

#check the totals
sum(mono.grad)
sum(Value)
```

### PCLM
The `pclm` option in `DemoTools` ungroups data assuming the counts are distributed Poisson over a smooth grid [@pascariu2018ungroup]. The grouped data can be efficiently ungrouped by using the penalized composite link model [@rizzi2015efficient]. This method is a flexible choice since the only assumption underlying the model is smoothness, which is usually met with demographic counts. For more details see @rizzi2015efficient and the `ungroup` package in R.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
pclm.grad <- graduate(Value, Age, method = "pclm")

single.age  <- names2age(pclm.grad)
plot(single.age, pclm.grad, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with pclm')

#check the totals
sum(pclm.grad)
sum(Value)
```


### Uniform
With the `uniform` option graduation is performed assuming that the counts are uniformly distributed across each age interval. It is also assumed that the initial age intervals are integers greater than or equal to one. This option simply splits aggregate counts in age-groups into single age groups.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
unif.grad <- graduate(Value, Age, method = "unif")

single.age  <- names2age(unif.grad)
plot(single.age, unif.grad, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with uniform distribution')

#check the totals
sum(unif.grad)
sum(Value)
```

### Comparison of all methods
```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
plot(single.age, beers.ordinary, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with multiple methods')

lines(single.age,beers.modified, col = "red")

lines(single.age,sprague, col = "blue")

lines(single.age,mono.grad, col = "green")

lines(single.age,pclm.grad, col = "pink")

lines(single.age,unif.grad, col = "magenta")

legend("topright",
       lty=1,
       col=c('black','red','blue','green','pink','magenta'),
       legend=c("Beers ordinary",
                "Beers modified",'Sprague','Monotone','PCLM',
                'Uniform'))
#check the totals
sum(beers.ordinary)
sum(beers.modified)
sum(Value)
```


### Using a standard
The function `rescaleAgeGroups` rescales a vector of counts in some given age-groups to approximate the counts into different age grouping. This method is appealing when, for example, the user wants to rescale single ages to sum to 5-year age groups. It is particularly useful in processes of making consistent different datasets to provide the same structure of data prior to construction of a lifetable for example.

To do: TR: rescaleAgeGroups() I'm, not sure. How about you get DemoData death or pop counts from an HMD country and in single ages from Jorge. Then rescale them to match HMD counts in 5-year age groups? Just an idea.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
AgeIntRandom <- c(1L, 5L, 2L, 2L, 4L, 4L, 1L, 2L, 3L, 4L, 3L, 3L, 3L, 3L, 5L)
AgeInt5      <- rep(5, 9)
original     <- runif(45, min = 0, max = 100)
pop1         <- groupAges(original, 0:45, AgeN = int2ageN(AgeIntRandom, FALSE))
pop2         <- groupAges(original, 0:45, AgeN = int2ageN(AgeInt5, FALSE))
# inflate (in this case) pop2
perturb      <- runif(length(pop2), min = 1.05, max = 1.2)

pop2         <- pop2 * perturb

# a recursively constrained solution
pop1resc <- rescaleAgeGroups(Value1 = pop1, 
					AgeInt1 = AgeIntRandom,
					Value2 = pop2, 
					AgeInt2 = AgeInt5, 
					splitfun = graduate_uniform, 
                 recursive = TRUE)
pop1resc

```

## Graduation as a light smoother
The graduation family functions `graduate` work strictly with data in 5-year age groups. Therefore, graduation can be seen as a light smoother when dealing with counts that are not grouped in 5-year age groups. In order to graduate counts that are not in 5-year age groups, an additional step to do this is necessary prior to graduation. For example, if counts are reported in single age groups, then before graduating the user needs to groups them in 5-year age groups, which can be interpreted as a constrained smooth. Compared with the `agesmth` family of functions, in this case there would not be transfers between 5-year age groups.


## References
