---
title: "Graduation with DemoTools"
author: "José Manuel Aburto, Ilya Kashnitsky, Monica Alexander, Jorge Cimentada, Tim Riffe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Graduation with DemoTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6, 
  fig.height = 6,
  fig.align = "center"
)

```

## How to use `DemoTools` to graduate counts in grouped ages

### What is graduation?
Graduation is a practice used to derive figures for _n_-year age groups, for example 5-year age groups from census data, that are corrected for net reporting error [@siegel2004methods]. Essentially, these methods fit different curves to the original  _n_-year and redistribute them into single-year values. These techniques are designed so that the sum of the interpolated single-year values is consistent with the total for the groups as a whole. Among the major graduation methods are the Sprague [@sprague1880explanation] and Beers [@beers1945modified] osculatory methods, monotone spline and the uniform distributions. More recently, the enalized composite link model (pclm) was proposed to ungroup coarsely aggregated data [@rizzi2015efficient]. All of these are implemented in `DemoTools`.

### Why graduate?
One of the main purposes of graduation is to refine the detail of available data in published statistics or surveys. This allows to estimate numbers of persons in single years of age from data originally in 5-year age groups. In practice, graduation of mortality curves has been very important to estimate reliable life tables, but also in fertility studies graduation is useful for the analysis of data from populations where demographic records are defective [@brass1960graduation].

### How is this different from smoothing?
The main difference between graduation and smoothing is that graduation redistributes the counts or events into single-year age groups with the constrain that they sum to the total.The selection of the method to be used depends mostly on the balance between smoothness and closeness of fit to the data. Therefore, the analyst must decide which weight should be given to these two opposing considerations. 

## Read in data
```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
library(DemoTools)

# 'pop1m_ind' available as package data
Value <- pop5_mat[,4]
Age   <- as.integer(rownames(pop5_mat))
  
plot(Age, Value, type = 'p',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Original 5-year age-group population')


```

## Graduation basic usage

## Methods choices

### Sprague
@sprague1880explanation suggested the fifth-difference method based on one equation that depends on two polynomials of the fourth degree forced to have a common ordinate, tangent and radius of curvature at $Y_{n+2}$ and at $Y_{n+3}$:

\begin{equation*}
y_{n+2+x} = \frac{(x+2)}{1!}\Delta y_n +
\frac{(x+2)(x+1)}{2!}\Delta y^2_n +
\frac{(x+2)(x+1)x}{3!}\Delta y^3_n +
\frac{(x+2)(x+1)(x-1)x}{4!}\Delta y^4_n +
\frac{x^3(x-1)(5x-7)}{4!}\Delta y^5_n 
\end{equation*}

There are six observations $Y_{n},Y_{n+1},Y_{n+2},Y_{n+3},Y_{n+4}$ and $Y_{n+5}$ involved in the leading differences, $\Delta y_{n},\dots,\Delta^5 y_n $. The interpolated points fall on curves that pass through the given points, and the interpolated subdivisions add up to the data in the original groups. To implement this method in `DemoTools` we use the function `graduate` with the option for method specified as 'sprague. 


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
beers.ordinary <- graduate(Value, Age, method = "beers(ord)")
beers.modified <- graduate(Value, Age, method = "beers(mod)")

single.age  <- names2age(beers.ordinary)
plot(single.age, beers.ordinary, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with Beers methods')
lines(single.age,beers.modified, col = "red")
legend("topright",lty=1,col=c(1,2),legend=c("ordinary (5-yr constrained)","modified (smoother)"))
#check the totals
sum(beers.ordinary)
sum(beers.modified)
sum(Value)
```
### Beers

Following a similar idea, Beers interpolated two overlapping curves minimizing the squares of the differences within the interpolation range [@beers1945six,@beers1945modified]. Specifically, Beers did this by minimizing fifth differences for a six-term formula, refered to as the 'Ordinary' Beers method [@beers1945six]. Subsequently, the ordinay formula was modified to relax the requirement that the given value be reproduced and yield smoother interpolated results, refered to as 'Modified' Beers method [@beers1945modified].

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
sprague <- graduate(Value, Age, method = "sprague")
plot(single.age, sprague, type = 'l',
     ylab = 'Population',
     xlab = 'Age',
     main = 'Graduation with Sprague method')

#check the totals
sum(sprague)
sum(Value)
```

### Monotone spline
Fritsch, F. N. and Carlson, R. E. (1980) Monotone piecewise cubic interpolation, SIAM Journal on Numerical Analysis 17, 238–246.
Hyman, J. M. (1983) Accurate monotonicity preserving cubic interpolation. SIAM J. Sci. Stat. Comput. 4, 645–654.

### Uniform

### Using a standard
Here the idea is to do the same thing by using `rescaleAgeGroups()` and a standard single age structure. One rescales the standard age structure to match the grouped age data in bins. YMMV.

## Graduation as a light smoother
light heaping can be grouped and then graduated, holding counts constrained.

## Graduation as a lifetable step
We could do abridged tables, which have their own consequential a(x) assumptions, or we could graduate to single ages and make a simpler lifetables. In this case the a(x) assumption is partly outsourced to the graduation and partly simplified by assuming midpoints (usually).

## References
