---
title: "Smoothing with DemoTools"
author: "Jose Manuel Aburto, Ilya Kashnitsky, Marius Pascariu, Tim Riffe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Smoothing with DemoTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## How to use `DemoTools` to smooth population counts

Smoothing, grouping or graduation of data over age is traditionally intended to have plausible/corrected estimates of population counts from census data. This procedures help to derive figures that are corrected primarly for net error by fitting different curves to the original 5 or 10-year totals, modifying the original counts [@siegel2004methods]. Several methods have been developed for this aim and the major graduation methods are included in `DemoTools`. Including the Carrier-Farrag [@carrier1959reduction], Arriaga [@arriaga1994population], Karup-King-Newton, United Stations [@united1955manual], Spencer [@spencer1987improvements] and Zelnik methods. Below we briefly give an overview of the method and apply them to the male Indian population in 1991.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
library(DemoTools)

# 'pop1m_ind' available as package data
Value <- pop1m_ind
Age   <- 0:(length(Value)-1)
  
plot(Age, Value/sum(Value), type = 'l',
     ylab = 'Proportion of population',
     xlab = 'Single age',
     main = 'Unsmoothed population') 

  
```

### Carrier-Farrag

This method considers the ratio, $K$, of the population in one quinary age-group to the next one [@carrier1959reduction]. If $v_0$ is a denary age-group, and $v_{-2}$ and $v_2$ are the preceding and succeeding age-groups, respectively, and if $K^4 = v_{-2}/v_2$. Then, the older quinary group $v_1$ can be estimated by $v_0/(1+K)$. This equation connects the population in two ten-year age groups separated by an interval of ten years. Therefore the value $K$ can be seen as the middle point between the two ten-year age groups. To run this method in `DemoTools` the function `afesmth` is used with the option 'Carrier-Farrag'. The figure below  shows the smoothed population by five-year age groups.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Carrier-Farrag", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Carrier-Farrag',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```

### Arriaga

Similarly to the previous method, when the 10-year age group to be separates is the central group of three, the following formulas are used in this method [@arriaga1968new]:

\begin{equation*}
_{5}P_{x+5} = \frac{-_{10}P_{x-10}+11 _{10}P_{x+10}+2 _{10}P_{x+10}}{24}
\end{equation*}
and
\begin{equation*}
_{5}P_{x} = _{10}P_{x} - _{5}P_{x+5}
\end{equation*}

Where: $_{5}P_{x+5}$ is the population between ages $x+5$ and $x+9$; $_{10}P_{x}$ is the population between ages $x$ and $x+9$; and $_{5}P_{x}$ is the population between ages $x+$ and $x+4$. When the 10-year age group to be separated is an extreme age group (the youngest or the oldest), the formulas are different. For the youngest age
group, the following formulas are used:


\begin{equation*}
_{5}P_{x+5} = \frac{8 _{10}P_{x}+ 5 _{10}P_{x+10} - _{10}P_{x+20}}{24}
\end{equation*}
and
\begin{equation*}
_{5}P_{x} = _{10}P_{x} - _{5}P_{x+5}
\end{equation*}

and for the last age group the coefficients are reversed:

\begin{equation*}
_{5}P_{x} = \frac{ -_{10}P_{x-20}+ 5 _{10}P_{x-10}+ 8 _{10}P_{x}}{24}.
\end{equation*}

To perform this model the option 'Arriaga' must be chosen in the 'agesmth' function.


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Arriaga", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Arriaga',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```

### Karup-King-Newton 

Following the same logic, the KKN method uses the following formulas:


\begin{equation*}
_{5}P_{x} = \frac{1}{2} _{10}P_{x} + \frac{1}{16} \big( _{10}P_{x-10} - _{10}P_{x+10} \big)
\end{equation*}
and
\begin{equation*}
_{5}P_{x+5} = _{10}P_{x} - _{5}P_{x}.
\end{equation*}

To implement this smoothing process select the `Karup-King-Newton` inthe `agesmth` function.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Karup-King-Newton", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Karup-King-Newton ',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```

### United Nations
The United Nations [@carrier1959reduction] developed the following formula to smooth population counts
\begin{equation}
_5\hat{P}_x = \frac{- _5P_{x-10} + 4\, _5P_{x-5} + 10\, _5P_{x} +4\, _5P_{x+5} - _5P_{x+10}}{16}
\end{equation}

where $_5\hat{P}_x$ represents the smoothed population between ages $x$ and $x+4$. This method can be applied in `DemoTools` with the function `united_nations_smth` as follows

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
 un_result <- united_nations_smth(Value = Value,Age = Age,OAG = T)
plot(seq(0,100,5),un_result/sum(un_result,na.rm = T), type= 'l',
    ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with UN Method',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
```

### Strong
The Strong formula adjusts proportionally the smoothed 10-year age groups to the census population in those ages, after this procedure the 10-year age groups can be subdivided into 5-year age groups [@arriaga1968new]

\begin{equation}
_{10}\hat{P}_x = \frac{_{10}P_{x-10} + 2\, _10P_{x} + _{10}P_{x+10}}{4}
\end{equation}

where $_{10}\hat{P}_x$ represents the smoothed poopulation ages $x$ to $x+9$. It is implemented in `DemoTools` as follows


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
 strong_result <- strong_smth(Value = Value,Age = Age,OAG = T)
plot(seq(0,100,5),strong_result/sum(strong_result,na.rm = T), type= 'l',
    ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Strong formula',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
```

### Zigzag

The implementatios of this methods assumes that persons incorrectly reported in peak age groups are evenly divided between the two adjacent age groups [@feeney2013]. It relies in minimizing a measure of "roughness":
consider the difference $R[i]$ between the number of persons in the $i$-th age group and the
average of the numbers in adjacent age groups, $R[i] = N[i] - (N[i -1] + N[i + 1])/2$.
If the distribution displays zigzag, the $R[i]$ will relatively large. If the distribution is smooth,
they will be relatively small. This suggests taking the sum of squares of these differences as
the measure of roughness [@feeney2013]. It can be implemented in `DemoTools` as follows


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
 zz_result <- zigzag_smth(Value, Age, OAG = TRUE, ageMin = 0, ageMax = 100)
plot(seq(0,100,5),strong_result/sum(strong_result,na.rm = T), type= 'l',
    ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Zig Zag formula',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
```


### Polynomial
A more general way is to smooth trhough linear models. Polynomial  fitting is used to mooth data over age or time fitting linear models. It can be tweaked by changing the degree and by either log or power transforming and can be used on any age groups, including irregularly spaced, single age, or 5-year age groups. It can be implemented in `DemoTools` with the function `agesmth1` and the option `poly` as follows

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
 poly_result <- agesmth1(Value, Age, method="poly", OAG = T)
plot(0:100,poly_result, type= 'l',
    ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Poly formula',xaxt='n')
points(0:100,Value)
axis(1, labels =seq(0,100,5), at =seq(0,100,5))
```

### LOESS
LOESS (locally weighted smoothing) helps to smooth data over age, preserving the open age group if necessary. It is a popular tool to create a smooth line through a timeplot or scatter plot. It can be tweaked by changing the degree and by either log or power transforming and can be used on any age groups, including irregularly spaced, single age, or 5-year age groups.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
 loess_result <- agesmth1(Value, Age, method="loess", OAG = T)
plot(0:100,loess_result, type= 'l',
    ylab = 'Proportion of population',
     xlab = 'Age-group',main = 'Smoothed population with LOESS formula',xaxt='n')
points(0:100,Value)
axis(1, labels =seq(0,100,5), at =seq(0,100,5))
```
##References

