---
title: "Smoothing with DemoTools"
author: "Jose Manuel Aburto, Ilya Kashnitsky, Marius Pascariu, Tim Riffe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## How to use `DemoTools` to smooth population counts

Smoothing, grouping or graduation of data over age is traditionally intended to have plausible/corrected estimates of population counts from census data. This procedures help to derive figures that are corrected primarly for net error by fitting different curves to the original 5 or 10-year totals, modifying the original counts [@siegel2004methods]. Several methods have been developed for this aim and the major graduation methods are included in `DemoTools`. Including the Carrier-Farrag [@carrier1959reduction], Arriaga [@arriaga1994population], Karup-King-Newton, United Stations [@united1955manual], Spencer [@spencer1987improvements] and Zelnik methods. Below we briefly give an overview of the method and apply them to the male Indian population in 1991.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
library(DemoTools)

# 'pop1m_ind' available as package data
Value <- pop1m_ind
Age   <- 0:(length(Value)-1)
  
plot(Age, Value/sum(Value), type = 'l',
     ylab = 'Porportion of population',
     xlab = 'Single age',
     main = 'Unsmoothed population') 

  
```

### Carrier-Farrag

This method considers the ratio, $K$, of the population in one quinary age-group to the next one [@carrier1959reduction]. If $v_0$ is a denary age-group, and $v_{-2}$ and $v_2$ are the preceding and succeeding age-groups, respectively, and if $K^4 = v_{-2}/v_2$. Then, the older quinary group $v_1$ can be estimated by $v_0/(1+K)$. This equation connects the population in two ten-year age groups separated by an interval of ten years. Therefore the value $K$ can be seen as the middle point between the two ten-year age groups. To run this method in `DemoTools` the function `afesmth` is used with the option 'Carrier-Farrag'. The figure below  shows the smoothed population by five-year age groups.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Carrier-Farrag", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Porportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Carrier-Farrag',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```

### Arriaga

Similarly to the previous method, when the 10-year age group to be separates is the central group of three, the following formulas are used in this method [@arriaga1968new]:

\begin{equation*}
_{5}P_{x+5} = \frac{-_{10}P_{x-10}+11 _{10}P_{x+10}+2 _{10}P_{x+10}}{24}
\end{equation*}
and
\begin{equation*}
_{5}P_{x} = _{10}P_{x} - _{5}P_{x+5}
\end{equation*}

Where: $_{5}P_{x+5}$ is the population between ages $x+5$ and $x+9$; $_{10}P_{x}$ is the population between ages $x$ and $x+9$; and $_{5}P_{x}$ is the population between ages $x+$ and $x+4$. When the 10-year age group to be separated is an extreme age group (the youngest or the oldest), the formulas are different. For the youngest age
group, the following formulas are used:


\begin{equation*}
_{5}P_{x+5} = \frac{8 _{10}P_{x}+ 5 _{10}P_{x+10} - _{10}P_{x+20}}{24}
\end{equation*}
and
\begin{equation*}
_{5}P_{x} = _{10}P_{x} - _{5}P_{x+5}
\end{equation*}

and for the last age group the coefficients are reversed:

\begin{equation*}
_{5}P_{x} = \frac{ -_{10}P_{x-20}+ 5 _{10}P_{x-10}+ 8 _{10}P_{x}}{24}.
\end{equation*}

To perform this model the option 'Arriaga' must be chosen in the 'agesmth' function.


```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Arriaga", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Porportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Arriaga',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```

### Karup-King-Newton 

Following the same logic, the KKN method uses the following formulas:


\begin{equation*}
_{5}P_{x} = \frac{1}{2} _{10}P_{x} + \frac{1}{16} \big( _{10}P_{x-10} - _{10}P_{x+10} \big)
\end{equation*}
and
\begin{equation*}
_{5}P_{x+5} = _{10}P_{x} - _{5}P_{x}.
\end{equation*}

To implement this smoothing process select the `Karup-King-Newton` inthe `agesmth` function.

```{r,out.width='\\textwidth', fig.width= 6, fig.height=6, fig.align='center'}
cf <- agesmth(Value = Value, 
	Age = Age, 
	method = "Karup-King-Newton", 
	OAG = TRUE)

plot(seq(0,100,5),cf/sum(cf), type= 'l',
     ylab = 'Porportion of population',
     xlab = 'Age-group',main = 'Smoothed population with Karup-King-Newton ',xaxt='n')
axis(1, labels = paste0(seq(0,100,5),'-',seq(4,104,5)), at =seq(0,100,5))
  
```



### United Nations

### Strong

### Zigzag

### Polynomial

### LOESS
<!-- LOESS (locally weighted smoothing) helps to smooth data over age, preserving the open age group if necessary. -->
<!-- #'  It is a popular tool to create a smooth line through a timeplot or scatter plot. -->
<!-- #'  Loess smoothness may be tweaked by specifying an optional \code{"span"} argument.  -->
<!-- #' Polynomial  fitting is used to mooth data over age or time fitting linear models.  -->
<!-- #' It can be tweaked by changing the degree and by either log or power transforming.  -->
<!-- #' The open age group can be kept as-is if desired by specifying \code{OAG = TRUE}.   -->
<!-- #' May be used on any age groups, including irregularly spaced, single age, or 5-year age groups.  -->
<!-- #' Predictions are only returned for the original age groups. -->

<!-- @examples -->
<!-- #' Age <- 0:99 -->
<!-- #'  #cols <- RColorBrewer::brewer.pal(7,"Reds")[3:7] -->
<!-- #'  cols <-  c("#FC9272", "#FB6A4A", "#EF3B2C", "#CB181D", "#99000D") -->
<!-- #' \dontrun{ -->
<!-- #'  plot(Age,pop1m_pasex) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="poly", OAG = FALSE),col=cols[1]) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="poly", degree = 3, OAG = FALSE), col = cols[2]) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="poly", degree = 3,  -->
<!-- #'        trans = "log", OAG = FALSE), col = cols[3]) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="poly", degree = 3,  -->
<!-- #'        trans = "power", pow = 2, OAG = FALSE), col = cols[4]) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="poly", degree = 4,  -->
<!-- #'        trans = "power", pow = 3, OAG = FALSE), col = cols[5]) -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="loess", OAG = FALSE), col = "royalblue") -->
<!-- #'  lines(Age, agesmth1(pop1m_pasex, Age, method="loess", span = 1, OAG = FALSE), col = "blue") -->


##References

