---
title: "Smoothing with DemoTools"
author: "Jose Manuel Aburto, Ilya Kashnitsky, Marius Pascariu, Tim Riffe"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes:
- \usepackage{amsmath}
- \usepackage{amssymb}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## How to use `DemoTools` to smooth population counts

Smoothing, grouping or graduation of data over age is traditionally intended to have plausible/corrected estimates of population counts from census data. This procedures help to derive figures that are corrected primarly for net error by fitting different curves to the original 5 or 10-year totals, modifying the original counts [@siegel2004methods]. Several methods have been developed for this aim and the major graduation methods are included in `DemoTools`. Including the Carrier-Farrag [@carrier1959reduction], Arriaga [@arriaga1994population], Karup-King-Newton, United Stations [@united1955manual], Spencer [@spencer1987improvements] and Zelnik methods. Below we briefly give an overview of the method and apply them to the male Indian population in 1991.

```{r}
library(DemoTools)

# 'pop1m_ind' available as package data
Value <- pop1m_ind
Age   <- 0:(length(Value)-1)
  
plot(Age, Value, type = 'l')  
  
```

### Carrier-Farrag

### Arriaga

### Karup-King-Newton

### United Nations

### Strong

### Zigzag



##References

